A Theory of Bidirectional Type Synthesis for Simple Types

### 2023-07-06 tex/sec5-synthesis.tex
[ tex ] Revise S3; exactly 4 pages

\section{Bidirectional Type Synthesis and Checking} \label{sec:type-synthesis}
This section is devoted to defining mode-correctness and deriving bidirectional type synthesis for any mode-correct bidirectional type system $(\Sigma, \Omega)$.
\Cref{subsec:mode-correctness} formalises the concept of mode-correctness and establishes the uniqueness of synthesised types.
This indicates that for a given raw term and context, any two synthesised types must be equal.
This characteristic is also used in the proof of the logical decidability of bidirectional type synthesis.
Later, in \Cref{subsec:trichotomy}, we introduce the decidability of bidirectional synthesis but postpone its proof until \Cref{subsec:bidirectional-synthesis-checking}.
At that point, we will combine mode preprocessing, soundness, and completeness from \Cref{sec:pre-synthesis} to show the trichotomy on raw terms as a simple corollary.
Note that the result is applicable to an arbitrary mode-correct bidirectional type system, rather than being restricted to the simply typed $\lambda$-calculus as we did in \Cref{sec:key-ideas}.

@@

\begin{corollary}[Trichotomy on Raw Terms]\label{cor:trichotomy}
  Let $(\Sigma, \Omega)$ be a mode-correct bidirectional type system.
  Then, for every context $\Gamma$ and raw term $\isTerm{t}$, exactly one of the following statements holds:
  \begin{enumerate}
    \item $\isTerm{t}$ is synthesising and the judgement $\Gamma |-_{\Sigma, \Omega} \isTerm{t} : A$ has a derivation for some type $A$.
    \item $\isTerm{t}$ is synthesising but the judgement $\Gamma |-_{\Sigma, \Omega} \isTerm{t} : A$ has no derivation for any type $A$.
    \item $\isTerm{t}$ is not synthesising.
  \end{enumerate}
\end{corollary}
\begin{proof}
  Combine  \Cref{thm:soundness,thm:completeness,thm:mode-preprocessing} with \Cref{thm:bidirectional-type-synthesis}.

>>>>

\section{Bidirectional Type Synthesis and Checking} \label{sec:type-synthesis}
This section is devoted to defining mode-correctness and deriving bidirectional type synthesis for any mode-correct bidirectional type system $(\Sigma, \Omega)$.
\Cref{sec:mode-correctness} formalises the concept of mode-correctness and establishes the uniqueness of synthesised types.
This indicates that for a given raw term and context, any two synthesised types must be equal.
This characteristic is also used in the proof of the logical decidability of bidirectional type synthesis.
Later, in \Cref{subsec:trichotomy}, we introduce the decidability of bidirectional synthesis but postpone its proof until \Cref{subsec:bidirectional-synthesis-checking}.
At that point, we will combine mode preprocessing, soundness, and completeness from \Cref{sec:pre-synthesis} to show the trichotomy on raw terms as a simple corollary.
Note that the result is applicable to an arbitrary mode-correct bidirectional type system, rather than being restricted to the simply typed $\lambda$-calculus as we did in \Cref{sec:key-ideas}.

@@

\begin{corollary}[Trichotomy on Raw Terms]\label{cor:trichotomy}
  Let $(\Sigma, \Omega)$ be a mode-correct bidirectional type system.
  Then, for every context $\Gamma$ and raw term $\isTerm{t}$, exactly one of the following statements holds:
  \begin{enumerate}
    \item $\isTerm{t}$ is synthesising and the judgement $\Gamma |-_{\Sigma, \Omega} \isTerm{t} : A$ has a derivation for some type $A$.
    \item $\isTerm{t}$ is synthesising but the judgement $\Gamma |-_{\Sigma, \Omega} \isTerm{t} : A$ has no derivation for any type $A$.
    \item $\isTerm{t}$ is not synthesising.
  \end{enumerate}
\end{corollary}
\begin{proof}
  Combine  \Cref{lem:soundness-completeness,thm:mode-preprocessing} with \Cref{thm:bidirectional-type-synthesis}.
